### 2021-02-26 rad-icfp-2021/intro.tex
cleaned up refs.bib
starting section 3

\subsection{Contribution}

@@

We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure.
We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
Let's have a look at what we obtain on a few examples.

\begin{example}[First-order term]
On simple first-order terms, the output code looks like SSA code \cite{cytron1989efficient}.
Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
        !w_2'(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

@@

In Section~\ref{sec:background} we recall reduminents of automatic differentiation, and in particular the idea of reverse-mode and the cheap gradient principle.
Next, in Section~\ref{sec:simplediff} we introduce the source language we want to differentiate. We also introduce our new intermediate representation UNF, and a simple reverse-mode macro on this representation.
In Section~\ref{sec:efficientrad} we translate back from UNF to a purely functional target language and show several general optimisations to make our reverse-mode transformation efficient.
In the next Section~\ref{sec:complexity} we show that the previous optimisations are sufficient to ensure our transformation has the right complexity.
Section~\ref{sec:correctness} shows that our construction is correct and we give standard semantics to our source and target languages. 
In Section~\ref{sec:generalisation} we show how to generalise the previous sections for a richer source language. 
Finally, related work and some limitations to our approach are presentend in Section~\ref{sec:conclusion}.

>>>>

\subsection{Smmary of contributions}

@@

We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure.
We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
Let us have a look at what we obtain on a few examples.

\begin{example}[First-order term]
On simple first-order terms, the output code looks like SSA code \cite{cytron1989efficient}.
Let us consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
        !w_2'(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

@@

In Section~\ref{sec:background} we recall rudiments of automatic differentiation, and in particular the idea of reverse-mode and the cheap gradient principle.
Next, in Section~\ref{sec:simplediff} we introduce the source language we want to differentiate. We also introduce our new intermediate representation UNF, and a simple reverse-mode macro on this representation.
In Section~\ref{sec:efficientrad} we translate back from UNF to a purely functional target language and show several general optimisations to make our reverse-mode transformation efficient.
In the next Section~\ref{sec:complexity} we show that the previous optimisations are sufficient to ensure our transformation has the right complexity.
Section~\ref{sec:correctness} shows that our construction is correct and we give standard semantics to our source and target languages. 
In Section~\ref{sec:generalisation} we show how to generalise the previous sections for a richer source language. 
Finally, related work and some limitations to our approach are presentend in Section~\ref{sec:conclusion}.

### 2021-02-26 rad-icfp-2021/simplediff.tex
adding some structure to the document
starting unf figures
outline paper in intro
starting section 3

\section{Simple reverse-mode differentiation}

@@

@@

@@

@@

>>>>

\section{Simple reverse-mode differentiation}
\label{sec:simplediff}

@@

We consider a standard call-by-value language. 
It is a first-order language with array constructs and a few typical second-order array operations. 
The types !T1,T2! and terms !e1,e2! are in Figure~\ref{fig:source_grammar}.

\input{figures/source_grammar}

The typing rules are in Figure~\ref{fig:source_typesystem}. We have included a minimal set of array operations for the sake of illustration, 
but it is not difficult to add further operations. For scalar operations, we assume given a set of operations. 
!op1! and !op2! denote respectively unary and binary smooth operations on reals. 
We use infix notation for binary operators.
Typical examples include !cos, exp, +, *!. 
For the sake of simplicity in the development, the function in the array operations !map, map2, foldl! are quite restricted with and free variables. 
We also restrict to arrays of reals.
The general case is also useful and presented in Section~\ref{sec:generalisation}.

\input{figures/source_type_system}

@@

Following the intuition higlighted in Section~\ref{subsec:insights}, we present a new language which we call unary form (UNF). 
It simply consists of a composition of unary operators. That said, we want to compile our source language to this intermediate representation, 
and we need to remember some information about the initial term.

The grammar of our source UNF is given in Figure~\ref{fig:unf_source_grammar}.

\input{figures/source_unf}

There are a few notable things in this syntax. 
Every term is indexed by a well-defined contex $\Gamma$ from the source language.
In addition, a variable has a integer index $1\leq i\leq n$ where $n$ is the size of the context $\Gamma$.
Every constant, unary, or binary operator is summarised as an $n$-ary operator !op!$_n$.
Sequential composition is denoted by !;! and !e1;e2! means that !e1! should be performed, and then !e2!.
Every array operator !map!, !map2!, !foldl! has an extra index which represents a well-formed term in the source language.

The types are the same as in the source language. 
A context $\Gamma=\{$!x1: T1,...,xn:Tn!$\}$ can always be seen as a type !T1!$\times$...$\times$!Tn!.
UNF is very syntactic and does not really use variables. 
In fact it consists only of constants and composition. 
For this reason, we denote contexts as types in the indices of the terms, and $\times$ is used for context extension.
But we do emphasize that we mean contexts and the actual variables are remembered and
this will be key when compiling to the target language after differentiation.

The typing rules are detailed in Figure~\ref{fig:source_unf_typesystem}.
By $\reals^{\times n}$ we mean the product $\reals\times...\times\reals$ of $n$ factors $\reals$.

\MH{need to explain that return values produce some kind of ANF and introduce new vars!}

\input{figures/source_unf_typesystem}

We give a translation from our source language to source UNF in Figure~\ref{fig:source_to_unf}.

\MH{explain kind of monadic ; need theorem saying this makes sense.}

\input{figures/source_to_unf}

@@

\input{figures/target_unf}

\input{figures/target_unf_typesystem}

@@

\input{figures/macro_diff}
